### 2025-11-04 Proposals/2025-NSF-EntropicGeometry/NarrativeBody.tex
Add citations and some missing text

While the structures of the state spaces of classical and quantum mechanics are different, their spaces of ensembles have a rather similar structure: they both consist of a bounded convex subset of a metrizable and second-countable locally convex topological vector space, where the metric can be derived from the entropy through the square root of the Jensen-Shannon divergence $JSD = S\left(\frac{1}{2} \ens[a] + \frac{1}{2} \ens[b]\right) - \left(\frac{1}{2}S(\ens[a]) + \frac{1}{2}S(\ens[b])\right)$~\cite{Jensen-ShannonOriginal}. The JSD is, in both cases, the square of a distance function and can be interpreted as how much the average entropy increases under an equal mixture of two ensembles. There are, therefore, three common structures at play: a topology (which physically represents experimental verifiability), a convex structure (which physically represents the ability to make statistical mixtures) and an entropic structure (which physically represents the variability of the elements within each ensemble). While the axioms for the topology and the convex structure are readily available, to our knowledge, the correct axioms for the entropic structure still need to be identified. Moreover, the interplay between the topology, the convex structure and the entropic structure opens additional mathematical questions which have not been investigated.

@@

Note that in all above three cases the definition of the entropy is tightly bound to the definition of the space. Therefore, the biggest problem is understanding how one can give a single characterization that is general enough to encompass the three cases, while constrained enough to recover the correct expression in each case. Fortunately, in our preliminary work we already have found the solution. In all cases, the entropy possesses an upper bound to its increase under convex combination: $S(p\ens[a] + \bar{p} \ens[b]) \leq I(p) + p S(\ens[a]) + \bar{p} S(\ens[b])$ where $\bar{p} = 1-p$ and $I(p) = -p \log p - \bar{p}\log \bar{p}$. This maximum increase is given precisely when the two ensembles are orthogonal with respect to the inner product of the respective space. In the classical case, this means that the measures have disjoint support; in the quantum case, it means that the eigenspaces of non-zero eigenvalues are orthogonal. In other words, characterizing the interaction of the entropy function with the topology and the convex structure is enough to characterize the entropic structure.  That is, the form of $I(p)$---the Shannon entropy formula---can be characterized, and since in both classical and quantum mechanics ensembles can always be decomposed into orthogonal components, the different entropy formulas can be recovered in each case.

@@

Given these very promising preliminary findings, the goal of the proposed research is to find a minimal set of axioms to characterize the entropic structure of physical theories. Minimal in the sense that, in the spirit of analogous work on the foundations of mathematics, like set theory or reverse mathematics (cite revmath book), the axioms can be shown to be satisfied by the different theories (which now become models of the larger theory) but also allow one to rederive the entropy formula for each specific theory. Let us, then, go through the current state of our preliminary work, and show the type of questions that the overall approach touches.

@@

The second axiom imposes a convex structure, which represents the ability to perform statistical mixtures. Given two ensembles $\ens[a], \ens[b] \in \Ens$, we can always choose a probability $p \in [0,1]$ and select $\ens[a]$ $p$ percent of the times and $\ens[b]$ $\bar{p} = 1-p$ percent of the times. This gives the ensemble space a convex structure as $p \ens[a] + \bar{p} \ens[b] \in \Ens$ for all $\ens[a], \ens[b] \in \Ens$ and $p \in [0,1]$. This insight is already used in the literature, particularly in Generalized Probabilistic Theories (GPTs); see e.g.~Ref.~\cite{plavala2023} (cite wilce et al). The main innovation in this respect is that we do not limit ourselves to finite combinations and we do not simply allow all infinite combinations (which can be shown to include physically untenable cases (cite our Hilber and possibly the paper that shows entropy not defined for all distributions)). Rather, we let the topology do the work. That is, infinite convex combinations exist only if they are the topological limit of finite convex combinations. On physical grounds, the mixing operation must be continuous in the natural topology defined by experimental verifiability, much like addition and scalar multiplication are continuous in a topological vector space, leading to the idea of a topological convex space.

@@

Another line of inquiry is the relationship between the entropic structure and a metric over the ensemble space. For example, we can define the mixing entropy as $MS(\ens[a], \ens[b]) = S\left(\frac{1}{2}\ens[a] + \frac{1}{2} \ens[b]\right) - \left(\frac{1}{2} S(\ens[a]) + \frac{1}{2} S(\ens[b])\right)$. If we express the entropy in bits, the mixing entropy satisfies the following:
\begin{itemize}
	\item non-negativity: $MS(\ens[a], \ens[b]) \geq 0$
	\item identity of indiscernibles: $MS(\ens[a], \ens[b]) = 0 \iff \ens[a]=\ens[b]$
	\item unit boundedness: $MS(\ens[a], \ens[b]) \leq 1$
	\item maximality of orthogonals: $MS(\ens[a], \ens[b]) = 1 \iff \ens[a] \ortho \ens[b]$
	\item symmetry: $MS(\ens[a], \ens[b]) = MS(\ens[b], \ens[a])$
\end{itemize}
It is almost a distance function, as it does not satisfy the triangle inequality, but it recovers the Jensen-Shannon divergence (JSD)~\cite{Jensen-ShannonOriginal} in both classical and quantum mechanics. While the JSD is the square of a distance function in both classical and quantum information theory (see e.g.~\cite{JSDProperties}), our current axioms do not guarantee $MS$ to have this property in general. An interesting question, then, is whether enough physical constraints exist for that to be the case. The $MS$ can be used to create open balls and show that the topology must be Hausdorff. It is not yet clear whether this topology is metrizable and whether it must coincide with the topology of the ensemble space.

A related line of questions relates the entropic structure of a physical theory to elements of differential geometry and the calculus of variations. We are aware that the current axioms do not guarantee the entropic open balls to be convex, even though they are in both the classical and quantum cases. We suspect the open balls to be convex in general, even though we have not yet found a physically satisfying premise. If the convexity of the open balls is established, then the topology generated by entropic open balls is locally convex, which gives us enough structure to define calculus in infinite-dimensional spaces. That is, the entropic structure would in fact be responsible for the ability to define derivatives in physical theories.

In the finite-dimensional case, and if the entropy is twice differentiable with respect to the convex coefficients, the strict concavity of the entropy means that the Hessian is negative definite, which allows us to define a metric tensor. Let $V_{\ens[a]}$ be the embedding vector space with $\ens[a] \in \Ens$ as origin. This will be a differentiable manifold. On a linear coordinate chart we can define $g_{\ens}( v, w ) = - \frac{\partial^2 S}{\partial \ens ^2}( v, w )$ where $v$ and $w$ are elements of the tangent space at $\ens \in \Ens \subset V_{\ens[a]}$. Since all Hessians are symmetric, $g_{\ens}$ will be bilinear, symmetric and positive definite: it is a metric tensor and $V_{\ens[a]}$ is a Hessian manifold (cite). In classical and quantum mechanics it recovers the Fisher information metric [CA: Add a ref, maybe to an information theory textbook, since having trouble sorting out in which of Fisher's works this metric is introduced]. The immediate goal, then, is to see how much of information geometry can be generalized to this case. If a locally convex topology can be established, the further goal is to extend this to the infinite-dimensional case.

@@

A possible fourth axiom may be needed to conclude all the requirements for physical theories. In both classical and quantum mechanics, there is a Lie algebra that characterizes the relationships between generators of time evolution. In classical mechanics, this is captured by the Poisson bracket, while in quantum mechanics it is captured by the commutator. The unsolved difficulty is that these structures are defined on the state space and not on the ensemble space. While there are attempts in this direction (cite Reginatto and Ariel), we have found no general solution in the literature. Since the set of the probability measures defined on an odd-dimensional manifold satisfies the previous axioms, at least in the current form, it may be that the previous three axioms are not sufficient since no symplectic structure can exist over odd-dimensional manifolds. However, there is an intimate relationship between this Lie-algebraic structure and the entropic structure, since in both cases, the Lie-algebraic structure describes the relationship between the generators of transformations that leave the geometric structure (i.e.~symplectic form and inner product respectively) unchanged.  This means, as outlined before, they must leave the entropic structure unchanged. That is, it describes the group of deterministic and reversible motions, those that leave the entropy unchanged~\cite{aop-HamConsInfoEnt,aop-HamQuantInfo,aop-book}. Part of the investigation to identify the correct specification of the entropic structure, then, will need to consider its impact on the definition of the Lie-algebraic structure. For example, in our reverse physics approach to classical mechanics~\cite{aop-book,Carcassi:2022bpm}, we were able to show that a symplectic structure is necessary and sufficient to define an entropy that is invariant under frame transformations. If such a  requirement could be specified in general, the principle of relativity at equal time could be shown to be equivalent to this structure.

\noindent
\textbf{Summary of proposed research program:} As we have seen, the program is rich and connects to many areas of physics, information theory and mathematics. To not get lost, let us reiterate the primary goals. We want to mathematically formalize the intuition that specifying the geometry in physics is equivalent to specifying the entropy. To do so, we need to give a precise, physically motivated, axiomatic definition of what an entropic structure is for a general physical theory. The fundamental grounding concept will be the notion of statistical ensemble, as we can argue that any physical theory must \emph{at least} provide us with a space of statistical ensembles. All axioms will be stated as properties of ensemble spaces. Then we must show that the entropic structure characterizes the geometry of the general space in a way that recovers the geometry of classical and quantum mechanics in these specific cases. While pursuing this work, we will gain insight on how to handle topological issues properly, and hopefully show that ensemble spaces must embed into a second-countable metrizable locally convex vector space, so that the notion of differentiability can be extended to the infinite-dimensional case. We will also keep an eye open to connections to measure theory and Lie algebra, so that more features of physical theories can be generalized in the future. \\

[Notes:
Be clearer about how far expect to get in the 3 yrs.
Papers: 
1. Recover Shannon entropy as the maximum entropy increase under mixture for all ensemble spaces [can already do]
2. Show that we can recover the geometric structure of classical and quantum mechanics from the entropy structure alone [can already do]
3. Show that we can generalize information geometry in finite dimensions [can already do]
4. Show how the entropic structure constrains the space of physical theories and rules out some of the GPTs [cal already do]
]

[Expected significance/benefits: Can have a common axiomatic foundation for all physical theories]

[Add somewhere: Regarding the relationship of this work to the present state of knowledge in the field, while there exists other work attempting to reconstruct quantum mechanics from physical principles [cite], we are not aware of any literature attempting to relate the geometric and entropic structures of quantum mechanics, much less in a general physical theory.]
]

@@

@@

\section{Results from prior NSF support}
\subsection{Intellectual merit of prior NSF support}
Aidala has been a successful steward of NSF funds since receiving a CAREER Award in 2015.  Her prior research awards have been through the Physics Division, Nuclear Physics - Experiment program: \emph{CAREER: Valence and sea quark dynamics at Fermilab} (Award Number 1452636, Jul 2015-Jun 2021, \$738,082), and \emph{Studying quantum chromodynamics at LHCb} (Award Number 2012926, Jul 2020-Jun 2023, \$697,969; renewal Award Number 2309922, Jul 2023-Jun 2026, \$736,406).  She is additionally a Co-PI on an NSF AccelNet award, \emph{AccelNet-Design: Inter-American Network of Networks on Quantum Chromodynamics Challenges} (Award Number 2114669, Sep 2021-Aug 2026, \$38,556 allocated to U-M), aimed at further developing collaboration in quantum chromodynamics research between the U.S.~and Latin America.  Her QCD research has focused on advancing the multidimensional quark-gluon structure of the proton and on studying the processes by which scattered quarks or gluons form new bound states (hadrons).  Her group's NSF-funded work includes results published in Nature~\cite{SeaQuest:2021zxb}, Physical Review Letters~\cite{LHCb:2022tbc,LHCb:2019qoc}, Physical Review D Letters~\cite{LHCb:2022rky}, Physical Review D~\cite{LHCb:2025mcq}, Physical Review C~\cite{SeaQuest:2022xdu}, Physics Letters B~\cite{LHCb:2025tvf}, and Computing and Software for Big Science~\cite{Aidala:2023dai}, among other journals~[cite other results--Ezra's, SeaQuest methods, SeaQuest instrumentation], and it includes one additional paper submitted for review~[cite Cynthia's, and possibly ``two additional papers'' if Kara's submitted].  Regarding the specific scientific advancements of some of the work, Aidala and her collaborators have studied antimatter in the proton and found surprising evidence that there is an excess of down ``flavored'' antiquarks relative to up flavored antiquarks in the proton for antiquarks carrying as much as 45\% of the proton's total momentum~\cite{SeaQuest:2021zxb,SeaQuest:2022xdu}.  Her group also led the first-ever measurements of identified hadron species production by high-energy light-mass quarks in proton-proton collisions, multidifferential in multiple kinematic variables simultaneously, showing that existing models poorly predict the relative production rates of some species~\cite{LHCb:2022rky}, and recently discovered striking evidence of suppressed gluon radiation from beauty (bottom) quarks due to their large mass~\cite{LHCb:2025mcq}. In the spirit of making research more open, her group has championed and led the open data efforts of the 1200-scientist LHCb collaboration~\cite{Aidala:2023dai}.  This proposal is the second time she is submitting to the Applied Mathematics program, based on the research direction in the mathematical foundations of physics that she initiated more recently, which has so far produced 15 peer-reviewed papers since 2018 as well as a technical book, as described above in the section on qualifications of the PI and research team.  \\

>>>>

While the structures of the state spaces of classical and quantum mechanics are different, their spaces of ensembles have a rather similar structure: they both consist of a bounded convex subset of a metrizable and second-countable locally convex topological vector space, where the metric can be derived from the entropy through the square root of the Jensen-Shannon divergence $JSD = S\left(\frac{1}{2} \ens[a] + \frac{1}{2} \ens[b]\right) - \left(\frac{1}{2}S(\ens[a]) + \frac{1}{2}S(\ens[b])\right)$~\cite{Jensen-ShannonOriginal,HayashiQuantumInfo,majtey2005jensen}. The JSD is, in both cases, the square of a distance function and can be interpreted as how much the average entropy increases under an equal mixture of two ensembles. There are, therefore, three common structures at play: a topology (which physically represents experimental verifiability), a convex structure (which physically represents the ability to make statistical mixtures) and an entropic structure (which physically represents the variability of the elements within each ensemble). While the axioms for the topology and the convex structure are readily available, to our knowledge, the correct axioms for the entropic structure still need to be identified. Moreover, the interplay between the topology, the convex structure and the entropic structure opens additional mathematical questions which have not been investigated.

@@

Note that in all above three cases the definition of the entropy is tightly bound to the definition of the space. Therefore, the biggest problem is understanding how one can give a single characterization that is general enough to encompass the three cases, while constrained enough to recover the correct expression in each case. Fortunately, in our preliminary work we already have found the solution. In all cases, the entropy possesses an upper bound to its increase under convex combination: $S(p\ens[a] + \bar{p} \ens[b]) \leq I(p) + p S(\ens[a]) + \bar{p} S(\ens[b])$ where $\bar{p} = 1-p$ and $I(p) = -p \log p - \bar{p}\log \bar{p}$. This maximum increase is given precisely when the two ensembles are orthogonal with respect to the inner product of the respective space. In the classical case, this means that the measures have disjoint support; in the quantum case, it means that the eigenspaces of non-zero eigenvalues are orthogonal. In other words, characterizing the interaction of the entropy function with the topology and the convex structure is enough to characterize the entropic structure.  That is, the form of $I(p)$---the Shannon entropy formula~\cite{Shannon}---can be characterized, and since in both classical and quantum mechanics ensembles can always be decomposed into orthogonal components, the different entropy formulas can be recovered in each case.

@@

Given these very promising preliminary findings, the goal of the proposed research is to find a minimal set of axioms to characterize the entropic structure of physical theories. Minimal in the sense that, in the spirit of analogous work on the foundations of mathematics, like set theory or reverse mathematics~\cite{RevMathBook}, the axioms can be shown to be satisfied by the different theories (which now become models of the larger theory) but also allow one to rederive the entropy formula for each specific theory. Let us, then, go through the current state of our preliminary work, and show the type of questions that the overall approach touches.

@@

The second axiom imposes a convex structure, which represents the ability to perform statistical mixtures. Given two ensembles $\ens[a], \ens[b] \in \Ens$, we can always choose a probability $p \in [0,1]$ and select $\ens[a]$ $p$ percent of the times and $\ens[b]$ $\bar{p} = 1-p$ percent of the times. This gives the ensemble space a convex structure as $p \ens[a] + \bar{p} \ens[b] \in \Ens$ for all $\ens[a], \ens[b] \in \Ens$ and $p \in [0,1]$. This insight is already used in the literature, particularly in Generalized Probabilistic Theories (GPTs); see e.g.~Refs.~\cite{plavala2023,wilce2025generalizedprobabilitytheorynotes,DArianoBook} \textcolor{blue}{Anything else to cite?}. The main innovation in this respect is that we do not limit ourselves to finite combinations and we do not simply allow all infinite combinations (which can be shown to include physically untenable cases~\cite{baccetti2013infinite,aop-unphysHilbert}). Rather, we let the topology do the work. That is, infinite convex combinations exist only if they are the topological limit of finite convex combinations. On physical grounds, the mixing operation must be continuous in the natural topology defined by experimental verifiability, much like addition and scalar multiplication are continuous in a topological vector space, leading to the idea of a topological convex space.

@@

Another line of inquiry is the relationship between the entropic structure and a metric over the ensemble space. For example, we can define the mixing entropy as $MS(\ens[a], \ens[b]) = S\left(\frac{1}{2}\ens[a] + \frac{1}{2} \ens[b]\right) - \left(\frac{1}{2} S(\ens[a]) + \frac{1}{2} S(\ens[b])\right)$. If we express the entropy in bits, the mixing entropy satisfies the following:
\begin{itemize}
	\item non-negativity: $MS(\ens[a], \ens[b]) \geq 0$
	\item identity of indiscernibles: $MS(\ens[a], \ens[b]) = 0 \iff \ens[a]=\ens[b]$
	\item unit boundedness: $MS(\ens[a], \ens[b]) \leq 1$
	\item maximality of orthogonals: $MS(\ens[a], \ens[b]) = 1 \iff \ens[a] \ortho \ens[b]$
	\item symmetry: $MS(\ens[a], \ens[b]) = MS(\ens[b], \ens[a])$
\end{itemize}
It is almost a distance function, as it does not satisfy the triangle inequality, but it recovers the Jensen-Shannon divergence (JSD)~\cite{Jensen-ShannonOriginal} in both classical and quantum mechanics. While the JSD is the square of a distance function in both classical and quantum information theory (see e.g.~\cite{JSDProperties,HayashiQuantumInfo}), our current axioms do not guarantee $MS$ to have this property in general. An interesting question, then, is whether enough physical constraints exist for that to be the case. The $MS$ can be used to create open balls and show that the topology must be Hausdorff. It is not yet clear whether this topology is metrizable and whether it must coincide with the topology of the ensemble space.

A related line of questions relates the entropic structure of a physical theory to elements of differential geometry and the calculus of variations. We are aware that the current axioms do not guarantee the entropic open balls to be convex, even though they are convex in both the classical and quantum cases. We suspect the open balls to be convex in general, even though we have not yet found a physically satisfying premise. If the convexity of the open balls is established, then the topology generated by entropic open balls is locally convex, which gives us enough structure to define calculus in infinite-dimensional spaces. That is, the entropic structure would in fact be responsible for the ability to define derivatives in physical theories.

In the finite-dimensional case, and if the entropy is twice differentiable with respect to the convex coefficients, the strict concavity of the entropy means that the Hessian is negative definite, which allows us to define a metric tensor. Let $V_{\ens[a]}$ be the embedding vector space with $\ens[a] \in \Ens$ as origin. This will be a differentiable manifold. On a linear coordinate chart we can define $g_{\ens}( v, w ) = - \frac{\partial^2 S}{\partial \ens ^2}( v, w )$ where $v$ and $w$ are elements of the tangent space at $\ens \in \Ens \subset V_{\ens[a]}$. Since all Hessians are symmetric, $g_{\ens}$ will be bilinear, symmetric and positive definite: it is a metric tensor and $V_{\ens[a]}$ is a Hessian manifold~\cite{hessianmanifoldsShima1997,hessianmanifoldsliu2025}. In classical and quantum mechanics it recovers the Fisher information metric (see e.g.~Refs.~\cite{facchi2010,AmariInfoGeom}). The immediate goal, then, is to see how much of information geometry can be generalized to this case. If a locally convex topology can be established, the further goal is to extend this to the infinite-dimensional case.

@@

A possible fourth axiom may be needed to conclude all the requirements for physical theories. In both classical and quantum mechanics, there is a Lie algebra that characterizes the relationships between generators of time evolution. In classical mechanics, this is captured by the Poisson bracket, while in quantum mechanics it is captured by the commutator. The unsolved difficulty is that these structures are defined on the state space and not on the ensemble space. While there are attempts in this direction~\cite{caticha2008,HallReginattoBook}, we have found no general solution in the literature. Since the set of the probability measures defined on an odd-dimensional manifold satisfies the previous axioms, at least in the current form, it may be that the previous three axioms are not sufficient since no symplectic structure can exist over odd-dimensional manifolds. However, there is an intimate relationship between this Lie-algebraic structure and the entropic structure, since in both cases, the Lie-algebraic structure describes the relationship between the generators of transformations that leave the geometric structure (i.e.~symplectic form and inner product respectively) unchanged.  This means, as outlined before, they must leave the entropic structure unchanged. That is, it describes the group of deterministic and reversible motions, those that leave the entropy unchanged~\cite{aop-HamConsInfoEnt,aop-HamQuantInfo,aop-book}. Part of the investigation to identify the correct specification of the entropic structure, then, will need to consider its impact on the definition of the Lie-algebraic structure. For example, in our reverse physics approach to classical mechanics~\cite{aop-book,Carcassi:2022bpm}, we were able to show that a symplectic structure is necessary and sufficient to define an entropy that is invariant under frame transformations. If such a  requirement could be specified in general, the principle of relativity at equal time could be shown to be equivalent to this structure.

\noindent
\textbf{Expected outcomes and significance:}
During the three-year period of the proposed research, we expect to achieve and submit manuscripts on at least the following:
\begin{enumerate}
	\item Demonstrate that the Shannon entropy can be recovered as the maximum entropy increase under mixture for all ensemble spaces, and that the Shannon, Gibbs and von Neumann entropies can be recovered in the specific cases of discrete classical, continuous classical, and quantum mechanics
	\item Show that the geometric structures of classical and quantum mechanics can be recovered from the entropic structure alone 
	\item Demonstrate that information geometry can be generalized to the ensemble space of any physical theory of finite dimensions 
	\item Show that the entropic structure imposes constraints on the space of possible physical theories and rules out some Generalized Probabilistic Theories
\end{enumerate}
Additional possible outcomes during or subsequent to the period of proposed activities depending on the rate of advancement include \textcolor{blue}{[list near-term possible outcomes.  Then discuss possible longer-term future developments stemming from this work]}

Near-term possible: 
Find what results from information geometry and information theory could be generalized to ensemble spaces. \textcolor{blue}{Naturally lends itself to various self-contained contributions by the grad student, undergrads, or individuals outside the project}

Determine whether the convexity of the mixing entropy can be physically motivated and whether that guarantees a continuous embedding into a locally convex topological vector space.

Generalize information geometry to the infinite-dimensional case---would require a locally convex topology to be able to define variations.

Determine whether the ensemble spaces can be in general proven to be metrizable and characterize the connection between the metric and the entropy.

Generalize classical mechanics and classical field theory.

Find a suitable function space that can be used as a model for the case of classical field theories.

The primary significance of the proposed research is that it will \textcolor{blue}{[has the potential to?]} provide a common axiomatic foundation for \emph{all} physical theories.  Such an axiomatic foundation common to all physical theories will serve not only to better ground current theories and elucidate their relationships to one another, but also to constrain potential future physical theories, and perhaps point to assumptions or structures in current theories that hold promise for modification in order to develop new theories.

[We don't know where the mathematical structures in physics come from.  Therefore, when people try to generalize them, they do not know in which direction they can be generalized.  What we're doing is instead finding axioms from physical requirements---things the physics must at least give you.  The axiomatics are justified from the physics.  We're doing the due diligence to check that the axioms can be justified from the physics and they're independent of each other.  So when we're doing the generalization, the generalization is better constrained by the physics.]

Regarding the relationship of this work to the present state of knowledge in the field, while there exists other work attempting to reconstruct quantum mechanics from physical principles, for example in Refs.~\textcolor{blue}{[cite Hardy, Ariel, Pavia, see 2:36 into talk by Nicolas Medina Sanchez (Vienna) at Perimeter], we are not aware of any literature attempting to relate the geometric and entropic structures of quantum mechanics, much less in a general physical theory. One sentence about general approach(es) of those works?  There are people trying to reconstruct QM from physical premises, but these works are typically on finite dimensions and do not generalize to CM, and definitely don't generalize to classical or quantum field theories.}

\noindent
\textbf{Summary of proposed research program:} As we have seen, the program is rich and connects to many areas of physics, information theory and mathematics. Let us reiterate the primary goals. We want to mathematically formalize the intuition that specifying the geometry in physics is equivalent to specifying the entropy. To do so, we need to give a precise, physically motivated, axiomatic definition of what an entropic structure is for a general physical theory. The fundamental grounding concept will be the notion of statistical ensemble, as we can argue that any physical theory must \emph{at least} provide us with a space of statistical ensembles. All axioms will be stated as properties of ensemble spaces. Then we must show that the entropic structure characterizes the geometry of the general space in a way that recovers the geometry of classical and quantum mechanics in these specific cases. While pursuing this work, we will gain insight on how to handle topological issues properly, and hopefully show that ensemble spaces must embed into a second-countable metrizable locally convex vector space, so that the notion of differentiability can be extended to the infinite-dimensional case. We will also keep an eye open to connections to measure theory and Lie algebra, so that more features of physical theories can be generalized in the future. \\

@@

\textcolor{blue}{Mention that we expect to include undergrads}

@@

\section{Results from prior NSF support}
\subsection{Intellectual merit of prior NSF support}
Aidala has been a successful steward of NSF funds since receiving a CAREER Award in 2015.  Her prior research awards have been through the Physics Division, Nuclear Physics - Experiment program: \emph{CAREER: Valence and sea quark dynamics at Fermilab} (Award Number 1452636, Jul 2015-Jun 2021, \$738,082), and \emph{Studying quantum chromodynamics at LHCb} (Award Number 2012926, Jul 2020-Jun 2023, \$697,969; renewal Award Number 2309922, Jul 2023-Jun 2026, \$736,406).  She is additionally a Co-PI on an NSF AccelNet award, \emph{AccelNet-Design: Inter-American Network of Networks on Quantum Chromodynamics Challenges} (Award Number 2114669, Sep 2021-Aug 2026, \$38,556 allocated to U-M), aimed at further developing collaboration in quantum chromodynamics research between the U.S.~and Latin America.  Her QCD research has focused on advancing the multidimensional quark-gluon structure of the proton and on studying the processes by which scattered quarks or gluons form new bound states (hadrons).  Her group's NSF-funded work includes results published in Nature~\cite{SeaQuest:2021zxb}, Physical Review Letters~\cite{LHCb:2022tbc,LHCb:2019qoc}, Physical Review D Letters~\cite{LHCb:2022rky}, Physical Review D~\cite{LHCb:2025mcq}, Physical Review C~\cite{SeaQuest:2022xdu}, Physics Letters B~\cite{LHCb:2025tvf,SeaQuest:2024qdw}, and Computing and Software for Big Science~\cite{Aidala:2023dai}, among other journals~\cite{SeaQuest:2023tcr,SeaQuest:2017kjt}, and it includes one additional paper submitted for review~\cite{LHCb:2025rxf}.  Regarding the specific scientific advancements of some of the work, Aidala and her collaborators have studied antimatter in the proton and found surprising evidence that there is an excess of down ``flavored'' antiquarks relative to up flavored antiquarks in the proton for antiquarks carrying as much as 45\% of the proton's total momentum~\cite{SeaQuest:2021zxb,SeaQuest:2022xdu}.  Her group also led the first-ever measurements of identified hadron species production by high-energy light-mass quarks in proton-proton collisions, multidifferential in multiple kinematic variables simultaneously, showing that existing models poorly predict the relative production rates of some species~\cite{LHCb:2022rky}, and recently discovered striking evidence of suppressed gluon radiation from beauty (bottom) quarks due to their large mass~\cite{LHCb:2025mcq}. In the spirit of making research more open, her group has championed and led the open data efforts of the 1200-scientist LHCb collaboration~\cite{Aidala:2023dai}.  This proposal is the second time she is submitting to the Applied Mathematics program, based on the research direction in the mathematical foundations of physics that she initiated more recently, which has so far produced 15 peer-reviewed papers since 2018 as well as a technical book, as described above in the section on qualifications of the PI and research team.  \\
